---
title: 'Module 5 Overview Assessment: GLMMs, Vegan, and GGplot'
author: "Saif Elsaady"
date: "2025-02-17"
output: html_document
---

```{r setup, include=FALSE}
library(readr)

# Set file path 
file_path <- "~/Downloads/Mouse_Survival.csv"

# Read the CSV file 
mouse_data <- read_csv(file_path)
```

## Question 1: Mouse Survival
### Part 1 - Determining the Distribution of Mouse Survival Data
The mouse survival data follows a binomial distribution because survival is a binary outcome (0 = No, 1 = Yes)
```{r survival_distribution, echo=TRUE, message=FALSE, warning=FALSE}
### Part 1 - Determining the Distribution of Mouse Survival Data
library(fitdistrplus)
library(ggplot2)
str(mouse_data)

mouse_data$Ground_Cover <- as.factor(mouse_data$Ground_Cover)  # Convert to factor
mouse_data$Hawks_Present <- as.factor(mouse_data$Hawks_Present)  # Convert to factor

# Check unique values in the Survival_Yes column
table(mouse_data$Survival_Yes)
hist(mouse_data$Survival_Yes)

# The mouse survival data follows a binomial distribution because survival is a binary outcome (0 = No, 1 = Yes)

```
## Question 1: Mouse Survival
### Part 2 - Statistical Test for Survival Variation
A logistic regression model was used to look at how survival varies with ground cover, hawk presence, and mouse age.
```{r survival_analysis, echo=TRUE, message=FALSE, warning=FALSE}
### Part 2 - Statistical Test for Survival Variation
library(ggplot2)
library(stats)

# Convert categorical variables to factors
mouse_data$Ground_Cover <- as.factor(mouse_data$Ground_Cover)
mouse_data$Hawks_Present <- as.factor(mouse_data$Hawks_Present)

# Fit a logistic regression model (GLM) to predict survival
survival_model <- glm(Survival_Yes ~ Ground_Cover + Hawks_Present + Mouse_Age, data = mouse_data, family = binomial)
summary(survival_model)

#A logistic regression model was used to look at how survival varies with ground cover, hawk presence, and mouse age.
```
## Question 1: Mouse Survival
### Part 3 - Significance of Variables and Interpretation
All variables in the model (Ground Cover, Hawks Presence, and Mouse Age) are statistically significant (p < 0.001). Mice in dense vegetation have 17 times higher odds of survival compared to those in bare soil, while some vegetation also improves survival with a 5.5 times increase in odds. The presence of hawks reduces survival probability by approximately 40%, making it a significant negative factor. Additionally, older mice have a lower chance of survival, with a 12% decrease in survival probability per unit increase in age. The model explains approximately 80% of the variance in survival (Residual Deviance: 11039/13755 ≈ 80%), indicating a strong but not perfect predictive ability.
  

## Question 2: ESCA Trees
### Part 1 - Subset Data for "Ulmus parvifolia"
```{r subset_esca_trees, echo=TRUE, message=FALSE, warning=FALSE}
### Part 1 - Subset Data for "Ulmus parvifolia"
library(dplyr)
file_path <- "~/Downloads/ESCA_Trees.csv"
esca_data <- read.csv(file_path)
# Check structure of dataset
str(esca_data)
# Subset data for species "Ulmus parvifolia"
ulmus_data <- esca_data %>% filter(vegetation_scientific_name == "Ulmus parvifolia")

# Display
head(ulmus_data)
```

## Question 2: ESCA Trees
### Part 2 - Determine Distribution of "Height Measured"
It follows a nomral distribution. 0.4732 of a P value is good enough.
```{r height_distribution, echo=TRUE, message=FALSE, warning=FALSE}
library(fitdistrplus)
library(dplyr)
library(ggplot2)
# Subset data for "Ulmus parvifolia"
ulmus_data <- esca_data %>% filter(vegetation_scientific_name == "Ulmus parvifolia")

#loads of missing values

# Plot histogram
ggplot(data = ulmus_data, aes(x = height_measured)) +
  geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Height Measured (Ulmus parvifolia)", 
       x = "Height Measured", 
       y = "Frequency") +
  theme_minimal()

hist(ulmus_data$height_measured)

# Perform Shapiro-Wilk test
shapiro_test <- shapiro.test(ulmus_data$height_measured)

# Print test results
print(shapiro_test)

```

## Question 2: ESCA Trees
### Part 3 - Comparing Height by Vegetation Shape and Classification Code
The linear model analysis indicates that height (height_measured) is significantly influenced by vegetation classification, with naturalized trees being shorter on average (p = 0.00889, estimate = -1.9071) and shade trees being taller (p = 0.02176, estimate = 2.9429). The upright shape classification shows a marginal effect (p = 0.06765), while foundation and street classifications are not significant, meaning they do not strongly impact height. The model explains 71% of the variation (R² = 0.7095), and the F-statistic (5.861, p = 0.005742) confirms its overall significance. This suggests that some vegetation classifications affect height, and that others do not play a meaningful role.
```{r height_comparison, echo=TRUE, message=FALSE, warning=FALSE}
# Linear model: Height ~ Vegetation Shape + Vegetation Classification
# Linear model: Height ~ Vegetation Shape + Vegetation Classification
UP_LM <- lm(height_measured ~ vegetation_shape_code + vegetation_classification_code, 
            data = ulmus_data)
# Display summary of the linear model
summary(UP_LM)
```

## Question 2: ESCA Trees
### Part 4 - Comparing Height Between Vegetation Shape and Classification Code Combinations
The ANOVA results indicate that vegetation classification code significantly affects height (p = 0.00627).Suggesting different vegetation classifications lead to significant height variations.
```{r height_posthoc, echo=TRUE, message=FALSE, warning=FALSE}
# ANOVA to compare height across vegetation shape and classification code combinations
height_anova <- aov(height_measured ~ vegetation_shape_code * vegetation_classification_code, data = ulmus_data)

# Display ANOVA results
summary(height_anova)
```

## Question 3: HMF Plant Composition – Comparing Maple Composition
### Part 1 - Subset Data for Adult Maple Trees
```{r subset_maple_trees, echo=TRUE, message=FALSE, warning=FALSE}
### Part 1 - Subset Data for Adult Maple Trees
# Load necessary libraries
library(dplyr)

# Load dataset
file_path <- "~/Downloads/HMF_Plant_Comp_Data.csv"
hmf_data <- read.csv(file_path)

# Subset to include only adult maple trees (Genus Acer, no seedlings)
maple_trees <- hmf_data %>%
  filter(grepl("^Acer\\.", Species) & !grepl("seedling", Species))

# Display first few rows
head(maple_trees)
```


## Question 3: HMF Plant Composition – Comparing Maple Composition
### Part 2 - Two-Way ANOVA: Count Variation by Species and Time Period
```{r maple_anova, echo=TRUE, message=FALSE, warning=FALSE}
### Part 2 - Two-Way ANOVA: Count Variation by Species and Time Period
library(dplyr)
maple_data <- hmf_data %>%
  filter(grepl("^Acer", Species, ignore.case = TRUE) & !grepl("seedling", Species, ignore.case = TRUE))

# Run Two-Way ANOVA on Count.Cover by Species and Time_Period
anova_model <- aov(Count.Cover ~ Species * Time_Period, data = maple_data)

# Display ANOVA summary
summary(anova_model)
```





## Question 3: HMF Plant Composition – Comparing Maple Composition
### Part 3 - Visualization: Red Maple Count in Time Period 1 vs. Time Period 4
```{r red_maple_plot, echo=TRUE, message=FALSE, warning=FALSE}
### Part 3 - Visualization: Red Maple Count in Time Period 1 vs. Time Period 4
library(ggplot2)
library(dplyr)
# Subset data for Red Maple (Acer rubrum) in Time Periods 1 and 4
red_maple_data <- hmf_data %>%
  filter(Species == "Acer.rubrum" & Time_Period %in% c("Period_1", "Period_4"))

# Create the figure
ggplot(red_maple_data, aes(x = Time_Period, y = Count.Cover, fill = Time_Period)) +
  geom_violin(alpha = 0.5) +   # Show distribution
  geom_boxplot(width = 0.2, outlier.shape = NA) +  # Add boxplot for clarity
  labs(title = "Red Maple (Acer rubrum) Count by Time Period",
       x = "Time Period",
       y = "Count Cover") +
  theme_minimal() +  # Clean theme for better readability
  theme(legend.position = "none", text = element_text(size = 14))
```








## Question 4: HMF Community Analysis
```{r hmf_nmds_analysis, echo=TRUE, message=FALSE, warning=FALSE}
library(vegan)
library(dplyr)
library(tidyr)
library(ggplot2)
#PART 1
# Load dataset
file_path <- "~/Downloads/HMF_Plant_Comp_Data.csv"
HMF_Trees <- read.csv(file_path)

# Restructure Data for NMDS
HMF_Wide <- HMF_Trees %>%
  pivot_wider(names_from = Species, values_from = Count.Cover, values_fill = list(Count.Cover = 0))

# Remove first three columns to keep only numeric species data
HMF_Wide_NMDS <- HMF_Wide[, -c(1,2,3)]

#PART 2
# Perform NMDS using only species abundance data
HMF_NMDS <- metaMDS(HMF_Wide_NMDS, k = 2, trymax = 1000)

# Plot the NMDS result
plot(HMF_NMDS)

# Display NMDS points
HMF_NMDS$points


#Part 3
# Extract metadata (Site, Time_Period) for plotting
HMF_Plotting <- HMF_Wide[, c(1:3)]  # Select first three columns (Site, Time_Period, etc.)

# Add NMDS coordinates to metadata
HMF_Plotting$NMDS_X <- HMF_NMDS$points[, 1]
HMF_Plotting$NMDS_Y <- HMF_NMDS$points[, 2]

# Create NMDS plot
print(ggplot(data = HMF_Plotting, aes(x = NMDS_X, y = NMDS_Y, col = Time_Period)) +
  geom_point(size = 3, alpha = 0.7) +
  stat_ellipse(aes(group = Time_Period), linewidth = 1) +
  labs(title = "NMDS Plot of Community Composition",
       subtitle = "Comparison Across Time Periods",
       x = "NMDS Axis 1",
       y = "NMDS Axis 2",
       color = "Time Period") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right",
        legend.title = element_text(face = "bold"),
        legend.key = element_rect(fill = "white", color = "gray"),
        panel.grid = element_blank()))
```    


